A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-06-21 paper/paper.Rnw
Clean out addressed comments

%\fix{This paragraph may not be necessary, depending on the level of detail we need to get into.}
%\mt{I think briefly explaining the pre-training data helps support the idea that the model can learn to classify like we do, but it can definitely be said more concisely.}
Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images and spanning 1,000 catetories, that is commonly used in computer vision tasks; these images provide both a wide variety of images for training complex image-classification models and a standardized way to compare performance across different models.  \mt{\sout{Images from ImageNet are hierarchically labeled, so an image of a dog may have labels of sky, grass, and trees, and the dog may be labeled as ``mammal'', ``dog'', ``retriever'', and ``golden retriever'' at increasing levels of label complexity. }}CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. \mt{Later in this paper, we will examine a selection of filters from a CNN trained on ImageNet that demonstrates the generalizability of pre-trained models.} \fix{Consider citing the figure number; e.g. Figure xxx shows a selection of filters from VGG16 which demonstrate the levels of detail in the initial layers of a model trained on generic data. The only issue is that this means figures are cited out of order... not sure whether that's kosher.} \mt{I think I used to have that, but then you said I hadn't introduced VGG16 yet so it didn't make sense here (which also totally makes sense). It's definitely a chicken and egg problem...}

%\fix{This section repeats bits from the last two paragraphs. Consider reordering things so that there is less duplication? What's the best way to do that coherently?}
%\mt{Note to Susan: Most of the following paragraph has been Frankensteined into the 'Transfer learning' paragraph. If it sounds okay there, we can delete this paragraph.}
%Each layer in the model structure is comprised of sets of weights and connections that have been optimized during training. \mt{moved to para 1}
%Transfer learning in neural network contexts involves freezing the weights of the pre-trained model base taken from a network trained on a more general set of images, so that during the training process, only the weights of the newly-added classifier are updated \citep{oquabLearningTransferringMidlevel2014}. \mt{citation used in para 1}
%In many cases, the entire model base is used with a new model head; the modularity of CNNs makes this process relatively simple, and allows researchers to leverage pre-trained networks when working with smaller, domain-specific sets of image data. \mt{redundant to above, not used}
%Transfer learning allows CNNs to be applied to smaller datasets of several thousand images, reduces the amount of computational time required to fit the model, and provides boosted performance compared to training a new model from scratch \citep{yosinskiHowTransferableAre2014}. It has been successfully applied in automatic classification of medical images \citep{shinDeepConvolutionalNeural2016} as well as in various applications of shoe forensics \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017}. \mt{moved to end of para 1}

@@

\fix{I don't really want to get into convolutional layers, but if we're going to mention them we probably need to say something about them. Thoughts?}
\mt{I only want to give enough detail to briefly explain 1) how filters work as feature detectors and 2) why filters trained on not-shoes will work well on shoes. I think that point is made strongly by showing hierarchy in the blocks. I thought that the phrasing below gave just enough of a background to mention that filters are structured in layers, but that might raise more questions than it answers---maybe we could gloss over the layer-structure of the blocks and just mention they contain filters?}
The base of VGG16 is made up of five blocks, which consist of two or three convolutional layers and one max pooling layer which aggregates the signal from the convolutional layers. Each convolutional layer is made up of hundreds of filters, such that each of the five convolutional blocks has between 128 and 1,536 filters. A filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \autoref{fig:vgg16layers} shows a selection of filters from VGG16, with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors, lines, corners, and blobs, while later filters detect more complex combinations of features. This aggregation of features through successive convolutional layers mirrors the process of complex feature detection in human vision. Any complex feature can be disassembled into a set of simple features; similarly, the successive sets of layers detect increasingly detailed features.

@@

\paragraph{Annotated Training Data}
\fix{The first sentence is a repeat from earlier in the paper, so we may want to move directly into the feature set we're using. Thoughts?}
\mt{\sout{Class characteristics are characteristics which can be used to exclude shoes from a match at a crime scene, but cannot be used for individualized matching because they are shared by many shoes (e.g., all shoes of a specific brand and size).}} \mt{Existing research indicates that a} sufficiently well-defined set of features can be used to separate shoes into make and model categories \citep{grossVariabilitySignificanceClass2013}; the set of features used in that study included circle/oval, crepe, herringbone, hexagon, parallel lines, logo/lettering/numbering, perimeter lugs, star, and other. After consulting with practitioners, we developed a set of categories suitable for automatic recognition by convolutional neural networks. These modifications were necessary because some of the definitions used in Gross et al. (2013) require contextual spatial information which is not preserved during labeling (for example, lugs are required to be on the perimeter of the shoe). \autoref{tab:class-char-examples} shows three examples of each class \mt{in our feature set}.

@@

\subsection{Model Accuracy}
%
% \mt{\textbf{OUTLINE OF RESULTS}
% \begin{itemize}
% \item ROC: shows model performance is generally good, gives broad comparison of class performances
% \item Examples: show what predictions look like, the model isn't perfect, and motivate that some confusions are systematic between specific classes
% \item Confusion Matrix: show patterns in misclassification, motivate diagnostics to explain the cause of common mistakes
% \item Two main reasons for error: image contrast, and overlapping features between classes
% \begin{itemize}
% \item Low contrast: the model can't see features, which affects training and prediction
% \item Overlapping features:
% \begin{itemize}
% \item Many classes share individual features (e.g., lines, angles, curves), so the model already has to integrate features and learn context (e.g., circle vs o)
% \item Some classes are also much more variable (see brand plots), which means the model has to widen its definitions of certain shapes, which makes confusion between classes worse
% \end{itemize}
% \end{itemize}
% \item Heatmaps: 1) verify that the model is using category-appropriate features to make decisions and 2) show examples of understandable mistakes
% \end{itemize}
% }

\paragraph{Overall Accuracy}

For each image in the test set, the ``true" labels are nine human-assigned labels of 0 or 1 corresponding to the presence or absence of each of the nine shape categories in the image. When the CNN predicts which shapes are in the image, however, it assigns a probability between 0 and 1 for each shape category, with a total of nine probabilities per test image. Thus, to determine the accuracy of the model predictions, we must select a threshold for each label to discretize the model predictions. If the threshold is too low, moderate probabilites will be more likely to exceed the threshold, which increases the true positive and false positive predictions in the test data. Conversely, if the threshold is too high, moderate probabilities will more often fail to meet the threshold, which increases the number of true and false negative predictions. An appropriate threshold must be chosen to produce a high true positive rate while also ideally keeping the false positive rate low. One way to set an optimal threshold is to utilize Receiver Operating Characteristic (ROC) curves, a type of diagnostic plot which compare the false positive rate to the true positive rate for a classification method. \autoref{fig:overall-roc} shows the ROC curve for our model across all classes, and \autoref{fig:class-roc} shows the curve for each class. ROC curves can be summerized by the Area Under the Curve (AUC), which quantifies the overall accuracy of the model. Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance.

@@

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are corectly identified by the model above EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. \fix{These are mostly new examples from our last meeting because many of the previous images were in the training set... Grrrrr..... I have two or three too many, not sure about color contrast within images, etc., feel free to delete a few or we can discuss them. [Expand on specific misclassifications more, after determining which examples to delete.]}
\fix{We need to make sure that the x axis labels aren't crammed together... Overall the examples are good, but I think I might get rid of the all-white DC image, and possibly the one right below it as well. We can discuss on Monday.}
<<eers, include = F>>=
eers <- purrr::map_dbl(aucs$data$eer, function(x){x$thresholds}) %>%
  setNames(classes)
@

@@

% \mt{Side note: I keep futzing with this section. I am going to forcibly move on and welcome Susan's feedback, and not allow myself to change any more of the consistency plots.} \svp{Good strategy. Biggest feedback is to change the text to actually match the image as it is currently, but I get why you stopped and I'm proud of you for noticing that it was a problem and finding a solution. :) Nice job.}
\mt{As the above section is right now, the text should match the images... I had some issues with image sorting, because Windows will put 11 before 2 but not before 02, so maybe Linux has a different behavior that's resorting the images? I also cache them to speed up recompiling, so maybe something's up there?}

@@

%\svp{Pros and cons of confusion matrix here? Maybe we just don't show the actual proportions on the graph? I'd like to discuss this more in person.}

@@

\fix{Choose three (or four) heatmaps:
\begin{itemize}
\item An image with multiple true shapes that the model clearly predicts well, to prove that it has learned the features we think it should
\item (Seychelles circle/text confusion)
\item An image with high quad/tri/poly confusion, but the model uses appropriate features to make educated guesses
\item An image with no true classes but would yield false positives (e.g., corks, puzzle pieces) to show that there will always be weird shapes that our classification can't be expected to cover, but our model will still behave reasonably enough when encountering these cases.
\end{itemize}}

\mt{The captions on the old heatmaps are wrong since we switched models. Pls ignore :)}

\mt{I added EER options to heatmaps, but there are still some not ideal bits (it makes the absence of border look strange), so we can decide how to improve it or whether to use it}

<<setup-heatmap>>=
imgs <- list.files(file.path(wd, "images", "heatmaps", "raw"), full.names = T)
@

@@

# # Turn all images in "heatmaps/raw" to top-3 heatmaps
# sapply(imgs, function(x) {
#   calc_heatmap(x, loaded_model) %>%
#     prune_heatmap(top_n = 3, sort = "value") %>%
#     create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
# })

calc_heatmap(imgs[which(grepl("Ugg-grey", imgs))], loaded_model) %>%
  prune_heatmap(top_n = 3, sort = "value") %>%
  create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
                   eer = eers)

calc_heatmap(imgs[which(grepl("text", imgs))], loaded_model) %>%
  prune_heatmap(top_n = 3, sort = "value") %>%
  create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
                   eer = eers)

calc_heatmap(imgs[which(grepl("polygon", imgs))], loaded_model) %>%
  prune_heatmap(top_n = 3, sort = "value") %>%
  create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
                   eer = eers)

calc_heatmap(imgs[which(grepl("corks", imgs))], loaded_model) %>%
  prune_heatmap(top_n = 3, sort = "value") %>%
  create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
                   eer = eers)
#
# calc_heatmap(imgs[which(grepl("chevron-circle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("chevron", "quad", "circle")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("triangle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("triangle", "quad", "chevron")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("star", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "bowtie", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("other", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "chevron", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)

>>>>

%\fix{This paragraph may not be necessary, depending on the level of detail we need to get into.}
%\mt{I think briefly explaining the pre-training data helps support the idea that the model can learn to classify like we do, but it can definitely be said more concisely.}
Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images \mt{that span} 1,000 catetories, that is commonly used in computer vision tasks\mt{\sout{; these images provide both a wide variety of images for training complex image-classification models and a standardized way to compare performance across different models}}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. In the next section, we will introduce the pre-trained network VGG16, pictured in \autoref{fig:VGG16-structure}, and examine a selection of its filters (shown in \autoref{vgg16layers}) that demonstrate the levels of detail in the initial layers of a model trained on generic data.

@@

\mt{The base of VGG16 is made up of five convolutional blocks, each of which contain between 128 and 1,536 filters.} A filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \autoref{fig:vgg16layers} shows a selection of filters from VGG16, with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors, lines, corners, and blobs, while later filters detect more complex combinations of features. This aggregation of features through successive convolutional layers mirrors the process of complex feature detection in human vision. Any complex feature can be disassembled into a set of simple features; similarly, the successive sets of layers detect increasingly detailed features.

@@

\paragraph{Annotated Training Data}
\mt{Existing research indicates that a} sufficiently well-defined set of features can be used to separate shoes into make and model categories \citep{grossVariabilitySignificanceClass2013}; the set of features used in that study included circle/oval, crepe, herringbone, hexagon, parallel lines, logo/lettering/numbering, perimeter lugs, star, and other. After consulting with practitioners, we developed a set of categories suitable for automatic recognition by convolutional neural networks. These modifications were necessary because some of the definitions used in Gross et al. (2013) require contextual spatial information which is not preserved during labeling (for example, lugs are required to be on the perimeter of the shoe). \autoref{tab:class-char-examples} shows three examples of each class \mt{in the developed feature set}.

@@

\subsection{Model Accuracy}

\paragraph{Overall Accuracy}
For each image in the test set, the ``true" labels are nine human-assigned labels of 0 or 1 corresponding to the presence or absence of each of the nine shape categories in the image. When the CNN predicts which shapes are in the image, however, it assigns a probability between 0 and 1 for each shape category, with a total of nine probabilities per test image. Thus, to determine the accuracy of the model predictions, we must select a threshold for each label to discretize the model predictions. If the threshold is too low, moderate probabilites will be more likely to exceed the threshold, which increases the true positive and false positive predictions in the test data. Conversely, if the threshold is too high, moderate probabilities will more often fail to meet the threshold, which increases the number of true and false negative predictions. An appropriate threshold must be chosen to produce a high true positive rate while also ideally keeping the false positive rate low. One way to set an optimal threshold is to utilize Receiver Operating Characteristic (ROC) curves, a type of diagnostic plot which compare the false positive rate to the true positive rate for a classification method. \autoref{fig:overall-roc} shows the ROC curve for our model across all classes, and \autoref{fig:class-roc} shows the curve for each class. ROC curves can be summerized by the Area Under the Curve (AUC), which quantifies the overall accuracy of the model. Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance.

@@

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are corectly identified by the model above EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. \fix{These are mostly new examples from our last meeting because many of the previous images were in the training set... Grrrrr..... I have two or three too many, not sure about color contrast within images, etc., feel free to delete a few or we can discuss them. [Expand on specific misclassifications more, after determining which examples to delete.]}

\fix{DC moved to heatmap, green block text could still be removed}

<<eers, include = F>>=
eers <- purrr::map_dbl(aucs$data$eer, function(x){x$thresholds}) %>%
  setNames(classes)
@

@@

% \fix{make sure text matches figure!}

@@

@@

\fix{Choose three (or four) heatmaps:
\begin{itemize}
\item An image with multiple true shapes that the model clearly predicts well, to prove that it has learned the features we think it should (probably an UGG logo)
\item (Seychelles circle/text confusion)
\item An image with high quad/tri/poly confusion, but the model uses appropriate features to make educated guesses
\item An image with no true classes but would yield false positives (e.g., corks, puzzle pieces) to show that there will always be weird shapes that our classification can't be expected to cover, but our model will still behave reasonably enough when encountering these cases.
\end{itemize}}

@@

# Turn all images in "heatmaps/raw" to top-4 heatmaps
imgs <- list.files(file.path(wd, "images", "heatmaps", "raw"), full.names = T)
sapply(imgs, function(x) {
  calc_heatmap(x, loaded_model) %>%
    prune_heatmap(top_n = 4, sort = "value") %>%
    create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
})

# calc_heatmap(imgs[which(grepl("Ugg-grey", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("text", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("polygon", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("corks", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("chevron-circle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("chevron", "quad", "circle")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("triangle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("triangle", "quad", "chevron")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("star", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "bowtie", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("other", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "chevron", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
